
# Descriptive and summary statistics


```{r}
funcdir <- "C:/Users/MK/Documents/pCloud_Sync_offline/Examples/R/Functions"
# funcdir <- "F:/Examples/R/Functions"
paste(funcdir, c("tbl_freq.R",
                 "tbl_str.R",
                 "tbl_sum.R",
                 "timer.R",
                 "keep_obj.R",
                 "list_to_array.R",
                 "print_docs.R"), sep = "/") %>%
   lapply(., source) %>%
   invisible
```


Let's start with the following modified *mtcars* dataset:
```{r echo = FALSE}
df <- mtcars %>%
   mutate(ncyl = as.integer(cyl),
          automat = ifelse(am == 0, TRUE, FALSE),
          sizegrp = ifelse(disp <= 120, "small",
                           ifelse( disp >= 326, "big", "medium"))) %>%
   select(mpg, wt, ncyl, automat, sizegrp) %>% 
   as_tibble

str(df)
```


## Summarizing numeric variables


### Summary statistics using base R functions

The *summary* function returns a simple table of summary statistics for all/selected variables:
```{r}
summary(object = df)
```

The *agregate* function returns a statistics (specified by *FUN* argument) for selected variables by selected categories/classes/factors.
```{r}
aggregate(x = df[c("mpg", "wt", "ncyl")],
          by = list(auto = df$automat, size = df$sizegrp),
          FUN = mean)

aggregate(x = df[c("mpg", "wt", "ncyl")],
          by = df[c("automat", "sizegrp")], # An alternative way to specify "by" groups
          FUN = mean)
```

The *aggregate* function can only return one statistic at a time. The following example shows how it could be used to calculate more than one statistic using *lapply*. One could also use it in a for-loop to achieve the same effect.
```{r}
myfct <- function(fct) { aggregate(x = df[c("mpg", "wt", "ncyl")],
                                   by = list(auto = df$automat, size = df$sizegrp),
                                   FUN = fct) }

stats <- c("mean", "sd") # Vector of statistics to be calculated
stats_list <- lapply(stats, myfct)
names(stats_list) <- stats
stats_list
```


### Collapse a data frame into summary statistics using *dplyr*

Collapse a data frame into another data frame (a table of sorts) of selected summary statistics for selected variables using the *summarize* (or *summarise*, both are possible) function
```{r}
df %>%
   summarize(
      count = n(),
      mpg_nonmiss = sum(!is.na(mpg)),
      mpg_mean = mean(mpg, na.rm = TRUE),
      mpg_sd = sd(mpg, na.rm = TRUE)
   )
```

More frequently, *summarize* is used together with *group_by*. In the example below we form groups and calculate selected statistics by those groups (collapse a data frame into a table of summary statistics by group).
```{r}
df %>%
   group_by(automat, sizegrp) %>%
   summarize(
      count = n(),
      mpg_nonmiss = sum(!is.na(mpg)),
      mpg_mean = mean(mpg, na.rm = TRUE),
      mpg_sd = sd(mpg, na.rm = TRUE)
   )
```

Here is a more involved example of a summary table. The table (again, technically a data frame) is created and reshaped into a more convenient format (see the section on reshaping with *tidyr* for more information).
```{r}
df %>%
   group_by(automat, sizegrp) %>%
   summarize(
      count = n(),
      mpg_nonmiss = sum(!is.na(mpg)),
      mpg_mean = mean(mpg, na.rm = TRUE),
      mpg_sd = sd(mpg, na.rm = TRUE),
      ncyl_nonmiss = sum(!is.na(mpg)),
      ncyl_mean = mean(mpg, na.rm = TRUE),
      ncyl_sd = sd(mpg, na.rm = TRUE)
   ) %>% 
   gather(key = KEY, value = VALUE, -c(automat, sizegrp, count)) %>%
   separate(KEY, into = c("variable", "KEY"), sep = "_") %>%
   spread(KEY, VALUE) %>% 
   select(variable, automat, sizegrp, everything()) %>% 
   arrange(variable, automat, sizegrp)
```


### Summary statistics using the *psych* package

The *psych* package provides the functions *describe* to generate lists of several summary statistics, In effect, this is similar to collapsing a data frame as shown above:
```{r}
df[,c("mpg", "wt")] %>% 
   describe %>%
   as_tibble
```

The *describeBy* version of the function generates summary statistics by categories from one or more variable. Note that the groups can be defined in more than one way (`df[c("automat", "sizegrp")]` or as `list(df$automat, df$sizegrp)`). It takes additional steps to unpack the result into a convenient data frame. Note that it is important to specify `mat = TRUE` as the output is a matrix, which is easier to deal with than with a list.
```{r}
df[,c("mpg", "wt")] %>%
   describeBy(group = df[c("automat", "sizegrp")], mat = TRUE) %>%
   rownames_to_column(var = "row_name") %>% 
   as_tibble
```

There are other R packages (e.g., *Hmisc*) that also provide similar functions to create summary statistics tables.


## Summarizing categorical variables

Let's start with a simple data frame that contains 3 categorical variables
```{r echo = FALSE}
df <- tibble(x = c("a","a","a","b","b","c"),
             y = c("d","e","d","e","d","e"),
             z = c("f","f","f","g","g","g"))
df
```


### Univariate & multidimensional tabulations using base R

Functions *table* and *prop.table* produce the category frequency (count) and proportion tables, respectively:
```{r}
table(df$x, useNA = "ifany") # Frequency table (counts)
table(df$x, useNA = "ifany") %>% prop.table() # Proportions
```

If 2 or more variables are specified in *table*, the result will be a cross-tabulation.
```{r}
table(df$x, df$y, useNA = "ifany") # Frequency cross-tabulation for 2 variables
table(df$x, df$y, useNA = "ifany") %>% prop.table() # Proportions
```

Row and column marginals can be obtained usint the *margin.table()* function.
```{r}
table(df$x, df$y, useNA = "ifany") %>% margin.table(1) # Row-sum marginals
table(df$x, df$y, useNA = "ifany") %>% margin.table(2) # Column-sum marginals
```

An alternative for creating cross-tabulations is the function *xtable*. One of the differences is that we need to specify variables into the cross-tab using the formula notation:
```{r}
xtabs(formula = ~ x + y, data = df)
```

Note that the tabs and cross-tabs above are class table. They can be coerced into a data frame. This generates a long-format distribution table. 
```{r}
table(df$x, df$y, df$z, useNA = "ifany") %>%
   as.data.frame() %>% 
   mutate(
      Perc = round(Freq/sum(Freq), 3)
   ) %>% 
   rename(
      x_cats = Var1,
      y_cats = Var2,
      z_cats = Var3
   )
```


### User-defined custom tabulation function *tbl_freq()*

Custom tabulation and cross-tab function called *tbl_freq* is a wrapper for the native R function *table()*. The function returns a data frame, but can only take one argument (one categorical variable), i.e., it can't do cross-tabulations.
```{r}
tbl_freq(df$x)
```


### Tabulations using the *CrossTable* function

The *CrossTable* function from the *gmodels* package produces a nice-format table with both cell counts and proportions as well as with row and column marginals. The shortcoming is that the function can only produce cross-tabulation for at most 2 variables.
```{r}
tlist <- CrossTable(df$x, df$y)
str(tlist)
```

The list that the function returns can be worked with further using the usual ways:
```{r}
tlist$t
tlist$prop.row
tlist$prop.col
tlist$prop.tbl
```

