
# DT package (DataTable.js library in R)

```{r warning = FALSE, message = FALSE, results = 'hide', echo = FALSE}
rm(list = ls())
options(DT.options = NULL) # Reset global options for 'DT::datatable()' to defaults
```

To illustrate its functionality, let us generate a data frame *df* that contains a variety of variable classes.
```{r echo = FALSE}
nobs <- 100

words <- c("plan","arc","futuristic","trucks","team","frame","odd","sail") %>% 
   gtools::permutations(n = length(.), r = 3, v = ., repeats.allowed = FALSE) %>% 
   apply(., MARGIN = 1, FUN = paste, collapse = " ")
   

df <- tibble(
   date          = as.Date(1:nobs, origin = "2018-05-15"),
   n_of_items    = runif(nobs, min = 1, max = 5.5) %>% round(0) %>% as.integer,
   class         = sample(c("c1", "c2", "c3"), nobs, replace = TRUE) %>% as.factor,
   subclass      = sample(paste0("subc", 1:4), nobs, replace = TRUE) %>% as.factor,
   best_in_class = runif(nobs, min = 0, max = 1) %>% round(0) %>% as.logical,
   amount_in_cad = round(runif(nobs, min = 500, max = 5000), 3),
   amount_in_usd = round(runif(nobs, min = 500, max = 5000), 3),
   growth_rate   = round(runif(nobs, min = -0.5, max = 2), 3),
   some_strings  = sample(words, nobs, replace = TRUE)
   )
```

```{r}
str(df)
```


## Initialization options

The default settings renders a data table as follows:
```{r}
options(DT.options = NULL) # Reset global options to defaults
datatable(df)
```

The function *datatable* from the *DT* package creates an HTML table following default settings for the format. However, the *DataTables.js* library has a large number of initialization options, which make it very flexible to customize tables. We can specify these options in an R list, and the *datatable* function will automatically convert them to JSON (needed by *DataTables.js*).

Let's define some custom initialization options: (1) display only the first 5 rows, (2) rename the *sSearch* element (the search/filter window) to *Filter:*, (3) enable case-insensitive search using  regular expressions, (4) order observations by the values of the 3rd (ascending) and 4th (descending) columns.
```{r}
custom_options <- list( # (1)
                        pageLength = 5,
                        # (2)
                        language = list(sSearch = "Filter:"),
                        # (3)
                        search = list(regex = TRUE, caseInsensitive = FALSE),
                        # (4)
                        order = list(list(3, "asc"), list(4, "desc")) )
```

The options can be specified either locally (for a single specific table, within one instance of the *datatable* function)
```{r}
datatable(df, options = custom_options ) # Single-table application
```
or globally for all data tables by including *options(DT. options = ...)* as a standalone line of code, or by including it into specs for any specific table.
```{r}
options(DT.options = custom_options ) # This will apply to all tables from this point on
```


## Additional options

We can change specific elements of the format by tweaking the various options in the *datatable* function. For example, below we (1) hide the row names, (2) change the format of the column names to uppercase and remove underscores (note that this is only a cosmetic change that doesn't change the underlying variable names), (3) add a filter to each column, and (5) make the table cells editable.
```{r}
dt_01 <- df %>%
   datatable(
      # (1)
      rownames = FALSE,
      # (2)
      colnames = gsub("_", " ", colnames(df)) %>% toupper,
      # (3)
      filter = list(position = "top"),
      # (4)
      editable = TRUE
      ); dt_01
```


### Format columns using helper functions

Several helper functions allows us to tweak the format of columns. For example, below we set certain numeric variables as (1) a specific currency (Canadian dollars using symbol C$), (2) percentages, or (3) dates. We can also change aspects of the column style using attributes in *formatStyle*
```{r}
# Define some custom colors

or1 <- rgb(218,  99,  93, maxColorValue = 255) # orange 1
gb1 <- rgb(212, 217, 222, maxColorValue = 255) # grey-blue 1
gb2 <- rgb(112, 128, 144, maxColorValue = 255) # grey-blue 2
gb3 <- rgb( 54,  69,  79, maxColorValue = 255) # grey-blue 3

dt_01 %>% 
   formatCurrency("amount_in_cad", currency = "C$", digits = 1, dec.mark = ".") %>% 
   formatPercentage("growth_rate", digits = 2) %>% 
   formatDate("date", method = "toDateString") %>% 
   formatStyle(columns = "growth_rate",
               color = "white",
               fontWeight = "bold",
               backgroundColor = gb3)
```


### Cells

```{r}
dt_01 %>%
   # Cell background color by class/category (2 or more categories)
   formatStyle("class",
               target = "cell",
               backgroundColor =styleEqual(c("c1", "c2", "c3"), c("none", gb1, gb2))) %>%
   # Cell and font colors based on the numeric values
   formatStyle(columns = c("amount_in_cad", "amount_in_usd"),
               color = styleInterval(c(1000, 3000), c("green", "red", "navy")),
               backgroundColor = styleInterval(mean(df$amount_in_cad), c(gb1, or1))) %>%
   # Bars inside cells corresponding to the numeric value
   formatStyle("n_of_items",
               background = styleColorBar(df$n_of_items, "lightblue"),
               backgroundSize = "95% 75%",
               backgroundRepeat = "no-repeat",
               backgroundPosition = "center" )
```


## The *DT* package OLD EXAMPLES, NEED TO BE SORTED OUT

```{r warning = FALSE, message = FALSE, results = 'hide', echo = FALSE}
rm(list = ls())
options(DT.options = NULL) # Reset global options for 'DT::datatable()' to defaults
```

Generate a small data frame to use in the examples.
```{r warning = FALSE, echo = FALSE}
set.seed(123)
n <- 15

df <- tibble(
   date = seq(as.Date('2018-05-15'), as.Date('2018-05-20'), 1) %>% sample(n, replace = TRUE),
   fact = c("alpha", "beta", "gamma") %>% sample(n, replace = TRUE) %>% as.factor(),
   dum = ifelse(rnorm(n) > 0, TRUE, FALSE),
   x = runif(n, -5, 5)*1000,
   y = runif(n, 0, 1),
   z = c(100, 500, 1000, 1250, 2725) %>% sample(n, replace = TRUE) %>% as.integer()
   ) %>%
   arrange(date)

df
```



```{r}
options(DT.options = NULL)

df %>%
   #
   datatable(
      #
      caption  = "Table 1: This is a caption for the table.",
      colnames = toupper(colnames(.)),
      rownames = FALSE,
      class = "cell-border stripe",
      filter = "top",
      #
      options = list( pageLength = 5, lengthMenu =c(5, 10, nrow(.)),
                      scrollX = FALSE, scrollY = FALSE,
                      autoWidth = TRUE,
                      columnDefs = list( list(targets = 0, width = "100px"),
                                         list(targets = c(3:5), width = "30px"),
                                         list(targets = c(1,3), searchable = FALSE)
                      )
      )
      #
   ) %>% 
   #
   formatDate(1, "toDateString") %>% 
   formatRound(c("x","y"), 3) %>% 
   formatCurrency("z", currency = "", interval = 3, mark = ",", digits = 0)
```







<!-- ```{r} -->
<!-- options(DT.options = NULL) -->

<!-- df %>% -->
<!--    datatable(  -->

<!--       caption  = "Table 1: This is a simple caption for the table.", -->
<!--       #colnames = toupper(colnames(.)), -->
<!--       rownames = FALSE, -->
<!--       class = "cell-border stripe", -->

<!--       options = list( -->
<!--          scrollX = TRUE, scrollY = FALSE, -->
<!--          pageLength = 5, lengthMenu =c(5, 10, 15, 20, 25, Inf), -->
<!--          autoWidth = TRUE -->
<!--          #columnDefs = list(list(width = "100px", targets = c(0,4))) -->
<!--       ) -->
<!--    ) %>%  -->
<!--    formatDate("date", "toDateString") %>%  -->
<!--    formatRound(c(3,8), 3) %>% -->
<!--    formatRound("z", 0) %>%  -->
<!--    formatCurrency("z", currency = "", interval = 3, mark = ",") -->
<!-- ``` -->


<!-- ```{r} -->
<!-- options(DT.options = NULL) -->

<!-- df %>% -->
<!--    datatable() %>%  -->
<!--    formatDate(1, "toDateString") %>%  -->
<!--    formatRound(c(3,8), 3) %>% -->
<!--    formatRound(c("z","newz"), 0) %>%  -->
<!--    formatCurrency("z", currency = "", interval = 3, mark = ",") -->
<!-- ``` -->
