
# Functional programming

```{r warning = FALSE, message = FALSE, results = 'hide', echo = FALSE}
rm(list = ls())
```

See <http://adv-r.had.co.nz/Functionals.html> for a more thorough treatment of functionals and functional programming.

The *apply* family contains higher-order functions that apply other functions to slices of data from matrices, arrays, lists and data frames. These functions allow crossing the data in a number of ways, and thus avoid explicit use of loop constructs. They act on an input list, matrix or array and apply a named function with one or several optional arguments.


## Mapping to vectors: *vapply*

Use *vapply* to map a function to each element of a given vector. Note that we have to specify which class of vector is to be returned (logical in the example below).
```{r}
vapply( X = c(-2, -1, 0, 1, 2),
        FUN = function(x) { x >= 0 },
        FUN.VALUE = logical(1) )
```


## Mapping to arrays and data frames: *apply*

We can map a function to one or more dimensions of a given array (matrices and data frames are considered special cases of arrays). In the following example we map a function to both rows and columns (i.e., every element) of a matrix by setting `MARGIN = c(1,2)` (note that dimension 1 is rows, 2 is columns).
```{r}
m <- matrix(-5:9, ncol = 3)
apply( X = m, FUN = function(x) { x >= 0 }, MARGIN = c(1,2) )
```

Some functions, such as the *sum* take vectors as arguments. This means we can apply them to a whole dimension (row or column) of a matrix or data frame.
```{r}
apply( X = m, FUN = sum, MARGIN = 2 )
```

A more interesting example: create a function that will normalize every observation so that their mean = 0 and sd = 1. The function is applied to every column (vector) of a given data frame or matrix.
```{r}
normalize <- function(x) { (x - mean(x)) / sd(x) }
df <- as_tibble(matrix(1:15, ncol = 3))
xn <- apply(X = df, FUN = normalize, MARGIN = 2)

rbind(Mean = apply(xn, FUN = mean, MARGIN = 2),
      `St. dev.` = apply(xn, FUN = sd, MARGIN = 2))
```


## Mapping to lists: *lapply*

The function *lapply* maps a function to every element of a given list. The result is another list, each element of which is the result of applying a given function.
```{r}
l <- list( x = c(1:3), y = c(T,T,F,F), z = letters[1:3])
lapply(X = l, FUN = class)
```

Similarly to *apply*, *lapply* can be used with a data frame as an argument, however, the returned object is a list. In the example below, note the use of *na.rm*, an additional argument for the *sum* function, and the use of *unlist* to change the returned list into a matrix.
```{r}
df <- tibble( x = 1:5, y = c(1, 2, NA, NA, 5)); df
lapply(X = df, FUN = sum) %>% unlist
lapply(X = df, FUN = sum, na.rm = TRUE) %>% unlist
```


## Mapping by groups: *aggregate*

The *aggregate* function splits the data into subsets, applies a function to each (typically a summary statistic one), and returns the result. Note that *tapply* seems to be an equivalent function but I haven't had a good experience with it. I the following example, we apply *sum* to two selected variables separately for each unique combination of the categories in *g1* and *g2* variables.
```{r}
df <- tibble( g1 = c("a","a","a","b","b","b","b"),
              g2 = c("c","c","d","c","d","c","d"),
              x = c(1:7),
              y = x*10 ); df

# unique(expand.grid(df$g1, df$g2)) # All unique groups

aggregate(x = df[c("x","y")], by = df[c("g1","g2")], FUN = sum)
```


## Mapping a function over multiple inputs: *Map*

Note *mapply* performs the same way as *Map* and, additionally, tries to simplify the returned object. It's better to use *Map* along with *unlist* and control the outcome. The following example shows how to make three random draws from a normal distribution. The draws differ in the number of observation, and in the mean and standard deviation of the distribution, i.e., the *rnorm* function is supplied with 3 vector arguments over which it needs to be mapped. The result is a list that contain 3 vectors.
```{r}
nobs <- c(5, 10, 15)
mu   <- c(-5, 0, 10)
sig  <- c(0.25, 1, 1.75)

Map(f = rnorm, n = nobs, mean = mu, sd = sig)

# When using an anonymous function
# Map(function(n, mean, sd) rnorm(n, mean, sd), n = nobs, mean = mu, sd = sig)

```


## Mapping multiple functions over a single input

The strategy is to create a single custom function that contains all the multiple functions we want to apply to the selected object. The example below uses a single user-defined function that collects several summary stats function. This function is then applied to the objects of choice.
```{r}
x <- runif(n = 900, min = 0, max = 10)
m <- matrix(x, ncol = 3)
df <- as_tibble(m) 

calculate_stats <- function(x) {
   
   stopifnot(is.numeric(x))
   
   mean_x   <- mean(x)
   sd_x     <- sd(x)
   min_x    <- min(x)
   max_x    <- max(x)
   
   round(rbind(mean_x, sd_x, min_x, max_x), 2)
}

calculate_stats(x) # Apply to a vector
apply(m, FUN = calculate_stats, MARGIN = 2) # Apply to matrix cols
apply(df, FUN = calculate_stats, MARGIN = 2) # Apply to df cols (result is a matrix)
lapply(df, FUN = calculate_stats) # Apply to df cols (result is a list)

# The list can be truned into a data frame
l <- lapply(df, FUN = calculate_stats)
do.call(cbind.data.frame, l)

```


