
# Interactive maps with *leaflet* (OLD VERSION)

```{r warning = FALSE, message = FALSE, results = 'hide', echo = FALSE}
rm(list = ls())

funcdir <- "C:/Users/MK/Documents/pCloud_Sync_offline/Examples/R/Functions"
# funcdir <- "F:/Examples/R/Functions"
paste(funcdir, c("tbl_freq.R",
                 "tbl_str.R",
                 "tbl_sum.R",
                 "timer.R",
                 "keep_obj.R",
                 "list_to_array.R",
                 "print_docs.R"), sep = "/") %>%
   lapply(., source) %>%
   invisible
```
```

Leaflet-related links:

* See <https://rstudio.github.io/leaflet/> for an example and help.
* For additional map providers, see <https://leaflet-extras.github.io/leaflet-providers/preview/>.
* See <https://www150.statcan.gc.ca/n1/en/type/reference#geographicfilesanddocumentation> for geography-related info from Statistics Canada.

Data links:

* Climate and weather in Canada: <http://climate.weather.gc.ca/index_e.html>

Use the built-in data frame *quakes* to generate a practice data frame:
```{r}
df <- quakes[1:100, ] %>%
   as_tibble %>% 
   add_count(stations) %>% rename(nquakes = n) %>% 
   mutate(cat_quakes = cut(nquakes,
                           labels = paste("freq", 1:5, sep = "_"),
                           breaks = c(-Inf, 3, c(4,6), 8 ,Inf),
                           ordered_results = TRUE),
          bad_quake = ifelse(nquakes > 5, "Y", "N"),
          stations = paste0("Station #", stations))

glimpse(df, width = 70)
```

Summary statistics:
```{r echo = FALSE}
# df %>% summary

df %>%
   select_if(is.numeric) %>%
   summarize_all(.funs = c(Mean = mean,
                           Median = median,
                           Min = min,
                           Max = max)) %>% 
   gather(key = STUB_SUFFIX, value = VALUE) %>% 
   separate(STUB_SUFFIX, into = c("STUB", "SUFFIX"), sep = "_") %>% 
   spread(SUFFIX, VALUE) %>% 
   mutate_at(.vars = vars(-STUB), .funs = function(x) {round(x,1)} ) %>% 
   rename(Variable = STUB) %>% 
   print.data.frame()

tbl_freq(df$cat_quakes) %>%
   arrange(desc(freq)) %>%
   print.data.frame(digits = 3)

tbl_freq(df$bad_quake) %>%
   arrange(desc(freq)) %>%
   print.data.frame(digits = 3)
```


## Base maps

There are well over 100 maps that can be used from the leaflet package. Their names can be obtained using *names(provider)* function; the maps can be then evoked using the function *addProviderTiles()*. In the example below, let's select all maps from a specific provider, *OpenStreetMap*:
```{r}
tibble(prvdr = names(providers)) %>% filter(str_detect(prvdr, "OpenStreetMap"))
```

Other maps possibly of interest are listed below. Note that these are to be used as arguments in the *addProviderTiles* function.
```{r}
# addProviderTiles("OpenStreetMap.HOT")
# addProviderTiles("CartoDB")
# addProviderTiles(providers$OpenTopoMap, group = "Topo")
# addProviderTiles(providers$Esri.WorldTopoMap, group = "Esri")
# addProviderTiles(providers$Stamen.Toner, group = "Toner")
```


## Selected options for the maps


### Zooming restrictions

Render a specific base map (black & white one from *Open Street Map*). Note that there are no custom options specified except the one for minimum and maximum zooming:
```{r}
leaflet( options = leafletOptions(minZoom = 0, maxZoom = 5) ) %>% 
   addProviderTiles("OpenStreetMap.BlackAndWhite")
```


### Setting the initial view

Now, render a map and fit it within specific bounds using *fitBounds* function. THe bounds are specified by providing the longitude and latitude coordinates for two corners that will define the "box".
```{r}
leaflet( ) %>% 
   addProviderTiles("OpenStreetMap.HOT") %>%
   fitBounds( lng1 = min(df$long), lat1 = min(df$lat),
              lng2 = max(df$long), lat2 = max(df$lat))
```

Instead of defining the bounds, we have an option to set the initial view at a specific location (defined by its longitude and latitude) and zoom.
```{r}
leaflet() %>%
   addProviderTiles(providers$Esri.WorldTopoMap, group = "Esri") %>%
   setView(lng = -76.395124, lat = 44.678639, zoom = 15)
```


## Putting data on maps

There are many graphic elements and layers that can be utilized to put data on the map. For more information see, e.g., the help page for the *addControl* function.


### Pop-up markers

See more examples at <https://rstudio.github.io/leaflet/markers.html>. In the following example, we will use the *addMarkers* function to:

* Place pin markers in the location (specified as the longitude and latitude) of the geo-stations that collected the *quakes* data.
* For each location generate a pop-up with the number of earthquakes counted for that location.
* Each location is a geo-station, so we label it with its station number.

```{r}
leaflet() %>%
   addProviderTiles("OpenStreetMap.HOT") %>%
   addMarkers( lng = df$long, lat = df$lat,
               popup = as.character(df$nquakes),
               label = as.character(df$stations) )
```






```{r}
leaflet() %>%
   addProviderTiles("OpenStreetMap.BlackAndWhite") %>%
   addCircleMarkers( lng = df$long, lat = df$lat,
                     radius = df$nquakes*1.5
   )
```


