# Machine Learning in R


```{r warning = FALSE, message = FALSE, results = 'hide', echo = FALSE}
rm(list = ls())

datadir <- "C:/Users/MK/Documents/pCloud_Sync_offline/Examples/R/Data"
funcdir <- "C:/Users/MK/Documents/pCloud_Sync_offline/Examples/R/Functions"

paste(funcdir, c("tbl_freq.R",
                 "tbl_str.R",
                 "tbl_sum.R",
                 "timer.R",
                 "keep_obj.R",
                 "list_to_array.R",
                 "print_docs.R"), sep = "/") %>%
   lapply(., source) %>%
   invisible
```

Note: Some of the ML-related stuff is in *p_train_test_models.Rmd*. Consider merging the contents into this file.


## Package *mlbench*: a collection of ML problems and datasets

The *mlbench* package contains a collection of artificial and real-world machine learning benchmark problems, including several data sets from the UCI repository. For the package documentation see <https://cran.r-project.org/web/packages/mlbench/mlbench.pdf>. For the U of California in Irvine (UCI) machine learning repo see <https://archive.ics.uci.edu/ml/index.php>.

```{r}
# This returns the list of all datasets in the package
data(package = "mlbench")

# This returns the matrix containing the datasets' names and short description
data(package = "mlbench")$results[ , c("Item","Title")]

# Load specific datasets from the package into the workspace
data(list = c("BostonHousing", "Glass", "Ozone"), package = "mlbench")

```
