
# Various specialized packages


## Package *alphavantager*

```{r warning = FALSE, message = FALSE, results = 'hide', echo = FALSE}
rm(list = ls())
```

*Alpha Vantage* is a free service that enables users to get real-time and historical financial data. To access the service one has to obtaine a free-of-charge API key. The following function allows to set the API key:
```{r}
av_api_key("ABNHXIMTX5J2FFU8")
print(av_api_key()) # Display the key currently in use
```

Retrieve trading data for SPY symbol (S&P500 ETF) from Alpha Vantage. Note that the *av_fun* option specifies whether the series is to be daily or intraday; the option *outputsize* allows us to retrieve either the last 100 observations (*compact*) or all of them (*full*); the option *interval* is self-explanatory:
```{r}
av_get( symbol     = "SPY",
        av_fun     = "TIME_SERIES_DAILY",
        outputsize = "compact")

av_get( symbol     = "SPY",
        av_fun     = "TIME_SERIES_INTRADAY",
        interval   = "15min",
        outputsize = "full")
```

