
# Working with time series


```{r warning = FALSE, message = FALSE, results = 'hide', echo = FALSE}
rm(list = ls())
```


## Working with series that are not equally spaced

We are taking a time series that is not equally spaced (some dates are missing) and make it into an equally-spaced one by filling NAs for those days. The starting point is a time series from which the NAs has been dropped, and the series has been ordered by the date variable *d*.
```{r echo = FALSE}
df <- tibble( d = seq(from = as.Date("2018-05-15"), length = 15, by = "days"),
              x = runif(n = 15, min = 85.0, max = 87.0) %>% round(1) ) %>%
   mutate_all(funs(replace(., row_number() %in% c(1, 3, 4, 7, 9:10), NA))) %>%
   mutate(d = sample(d)) %>% 
   na.omit %>%
   arrange(d)
```
```{r}
df
```


We'll take the date variable *d*, expand it by generating the missing dates, and then save it as a new data frame *new_d*. Then we'll merge the original data frame *df* and merge it with *new_d* by the variable *d*; as a consequence, *NA* will be 'imputed' for all other variables.
```{r}
new_d <- tibble( d = seq(from = min(df$d), to = max(df$d), by = "days") )
full_join(df, new_d, by = "d") %>% arrange(d)
```

Note that merging can also be done using the base R function *merge*.
```{r}
merge(new_d, df, by = "d", all = TRUE) %>% as_tibble
```